### 2018-05-25 submission/manuscript.Rmd
make changes to title and abstract

@@

@@

@@

@@

# High Fidelity DNA Polymerase Introduces Bias into 16S rRNA Gene Sequencing Results

@@

**Results.** When analyzing fecal samples we observed a different number of OTUs between HiFi DNA polymerases at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). Our analysis identified these polymerase dependent differences in the number of OTUs as early as 20 cycles in mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). Chimera prevalence varied by  polymerase and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs.

**Conclusions.** HiFi DNA polymerase dependent differences in the number of OTUs and chimera prevalence makes comparison across studies difficult. Care should be exercised when choosing the polymerase and number of cycles to be used in 16S rRNA gene sequencing studies.

>>>>

# Read in fecal count data
fecal_numOTU_data <- read_csv("../data/process/tables/fecal_sub_sample_1000_count_table.csv") %>% 
  mutate(taq = factor(taq, 
                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5")), 
         cycles = factor(cycles, 
                            levels = c("15x", "20x", "25x", "30x", "35x"), 
                            labels = c("15x", "20x", "25x", "30x", "35x"))) %>% 
  group_by(taq, cycles) %>% 
  summarise(taq_median = median(numOTUs, na.rm = T), 
            taq_min = min(numOTUs, na.rm = T), 
            taq_max = max(numOTUs, na.rm = T)) %>% 
  mutate(range_diff = taq_max - taq_min)

@@

# Read in the Mock count data
mock_num_otu_data <- read_csv("../data/process/tables/mock_sub_sample_1000_count_table.csv") %>% 
  mutate(taq = factor(taq, 
                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>% 
  group_by(taq, cycles) %>% 
  summarise(taq_median = median(numOTUs, na.rm = T), 
            taq_min = min(numOTUs, na.rm = T), 
            taq_max = max(numOTUs, na.rm = T)) %>% 
  mutate(range_diff = taq_max - taq_min)

ref_mock_otu_num <- read_tsv("../data/process/zymo_mock_only.shared") %>% 
  select(numOtus)

@@

# Read in actual error data
error_data <- read_csv("../data/process/tables/mock_chimera_error_summary.csv") %>% 
                       filter(total_seqs >= 1000)

error_summary <- error_data %>% 
  group_by(taq, cycles) %>% 
  summarise_at(vars(mean_error, chimera_prevalence, seq_error_prevalence), funs(median, min, max)) %>% 
  select(taq, cycles, contains("chimera"), contains("mean_error"), contains("seq_error"))

@@

# Read in the regression data
r2_table <- read_csv("../data/process/tables/r2_otu_vs_chimera_summary.csv")

@@

# Error Introduced into 16S rRNA Gene Sequencing Results Varies by High Fidelity DNA Polymerase Used

@@

**Results.** When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases at 35 cycles (Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Additionally, the median number of OTUs vaired by HiFi DNA polymerase used (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). Variations in the number of OTUs detected by DNA polymerase was observable as early as 20 cycles in mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). There also was a large range in the number of OTUs by HiFi DNA polymerase at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2 = `r max(pull(r2_table, r2))`) and Kappa having the worse (R^2 = `r min(pull(r2_table, r2))`).

**Conclusions.** Although the variation in the number of OTUs observed in fecal samples could be due to certain polymerases capturing the biological variability better than others, this is unlikely to be the main reason for this difference. In mock community samples, the divergence in the number of OTUs between the polymerases which can be explained by chimera prevalence suggests that this is the main reason for differences between the polymerases. Ultimately, this variation makes comparison across studies difficult and care should be exercised when choosing the polymerase and number of cycles to be used in 16S rRNA gene sequencing studies.

### 2018-05-25 submission/manuscript.tex
make changes to title and abstract

\section{High Fidelity DNA Polymerase Introduces Bias into 16S rRNA Gene
Sequencing
Results}\label{high-fidelity-dna-polymerase-introduces-bias-into-16s-rrna-gene-sequencing-results}

@@

\textbf{Results.} When analyzing fecal samples we observed a different
number of OTUs between HiFi DNA polymerases at 35 cycles (P-value
\textless{} 0.0001). Our analysis identified these polymerase dependent
differences in the number of OTUs as early as 20 cycles in mock
communities (P-value = 0.002). Chimera prevalence varied by polymerase
and this variation persisted after chimera removal using VSEARCH. We
also observed positive correlations between chimera prevalence and the
number of OTUs.

\textbf{Conclusions.} HiFi DNA polymerase dependent differences in the
number of OTUs and chimera prevalence makes comparison across studies
difficult. Care should be exercised when choosing the polymerase and
number of cycles to be used in 16S rRNA gene sequencing studies.

>>>>

\section{Error Introduced into 16S rRNA Gene Sequencing Results Varies
by High Fidelity DNA Polymerase
Used}\label{error-introduced-into-16s-rrna-gene-sequencing-results-varies-by-high-fidelity-dna-polymerase-used}

@@

\textbf{Results.} When analyzing fecal samples we observed that the
range in the number of OTUs detected was not consistent between HiFi DNA
polymerases at 35 cycles (Accuprime = 84 - 106 (min - max) versus
Phusion = 84 - 136). Additionally, the median number of OTUs vaired by
HiFi DNA polymerase used (P-value \textless{} 0.0001). Variations in the
number of OTUs detected by DNA polymerase was observable as early as 20
cycles in mock communities (P-value = 0.002). There also was a large
range in the number of OTUs by HiFi DNA polymerase at 35 cycles
(Accuprime = 15 - 20 versus Phusion = 14 - 73). Chimera prevalence in
mock communities varied by polymerase with differences being most
notable at 35 cycles (Kappa = 5.71\% (median) versus Platinum = 26.62\%)
and this variation persisted after chimera removal using VSEARCH. We
also observed positive correlations between chimera prevalence and the
number of OTUs with Platinum having the highest (R\^{}2 = 0.974) and
Kappa having the worse (R\^{}2 = 0.478).

\textbf{Conclusions.} Although the variation in the number of OTUs
observed in fecal samples could be due to certain polymerases capturing
the biological variability better than others, this is unlikely to be
the main reason for this difference. In mock community samples, the
divergence in the number of OTUs between the polymerases which can be
explained by chimera prevalence suggests that this is the main reason
for differences between the polymerases. Ultimately, this variation
makes comparison across studies difficult and care should be exercised
when choosing the polymerase and number of cycles to be used in 16S rRNA
gene sequencing studies.
